---
title: "Final Project"
author: "Simon Gao"
date: "5/14/2020"
output: pdf_document
---
I pledge my honor that I have abided by the Stevens Honor System.  

# Section 1: A Statistical Report  
## Executive Summary  
This report summarizes my findings on the potential differences in using different explanatory variables for simple linear regression models and multiple regression models. Specifically, I will discuss my findings in studying how 3 variables—Acetic, H2S, and Lactic—affect the Taste of cheese. I chose this topic because I was assigned it as a final project for my Intermdiate Statistics course. This report will contain the following: an explanation of the overall study, an explanation of the statistical methods used, a preliminary analysis of the data set, an anaylsis on the regression models, and a summary of the results.

## The Study  
As cheddar cheese matures, a variety of chemical processes take place. The taste of matured cheese is related to the concentration of several chemicals in the final product. In a study of cheddar cheese from the LaTrobe Valley of Victoria, Australia, samples of cheese were analyzed for their chemical composition and were subjected to taste tests. The variable “Case” is used to number the observations from 1 to 30. “Taste” is the response variable of interest. The taste scores were obtained by combining the scores from several tasters. Three of the chemicals whose concentrations were measured were acetic acid, hydrogen sulfide, and lactic acid. For acetic acid and hydrogen sulfide (natural) log transformations were taken. Thus, the explanatory variables are the transformed concentrations of acetic acid (“Acetic”) and hydrogen sulfide (“H2S”) and the untransformed concentration of lactic acid (“Lactic”). We will analysis a variety of linear regression models to determine which is the best predictory for Taste.  

## Statistical Methodology
My goal is to compare simple and multiple linear regression models using 3 explanatory variables—Acetic, H2S, and Lactic—and Taste as a reponse variable. In this report, I will create simple linear regression models (total of 3 models) for each explanatory variable and then pair them for multiple regression models (total of 3 models). There are a lot of statistical calculations for regression analysis but we will utilize the software "R" to perform these calculations for us.  

## Preliminary Analysis of the Data  
Before we construct linear models or perform regression analysis, it is essential for us to graphically check the relationships between variables and determine any deviations from normality for the variables. Preliminary analysis, using QQ plots, did not show any significant deviations from normality. However, the scatterplots of each pair of variables, 6 in total, all displayed a positive association betwn the pairs.   

## Regression Analysis  
First, we created linear models using the 3 explanatory variables separately as predictors for taste. The table below summarizes the resulting linear models and their respective equations.  

Linear Model | $F$  | $P-value$ | $R^2$  | $s$
------------ | ---- | --------- | ------ | -----
Taste-Acetic | 12.11| 0.002     | 30.2%  | 13.82
Taste-H2S    | 37.29| $1.37*10^-6$| 57.1%| 10.83
Taste-Lactic | 27.55| $1.41*10^-5$| 49.6%| 11.75  

$\hat{Taste}$ = -61.5 + 15.65 * $Acetic$  
$\hat{Taste}$ = -9.787 + 5.776 * $H2S$  
$\hat{Taste}$ = -29.86 + 37.72 * $Lactic$  

Out of all the simple linear regression models, the one with H2S seems to be the best one as its adjusted $R^2$ value is the highest at 57.1%. This implies that the linear model accounted for 57.1% of variation in Taste. Moreover, the model with Acetic as a predictor seemed to be the worst model out of the 3 simple linear regression models as it had the lowest adjusted $R^2$ value at 30.2%. In all 3 models, their residual plots did not have any striking deviations from a normal distribution. Additionally, the p-values for coefficients H2S and Lactic are significantly less than that of coefficient Acetic. Although the Acetic model shows a statistically significant relationship to Taste, it is not as statistically significant as the relationship of H2S and Lactic to Taste.  
  
As for the remaining 3 regression models, we created multiple regression models (see table below).  

Linear Model | $F$  | $P-value$ | $R^2$  | $s$
------------ | ---- | --------- | ------ | -----
Taste-Acetic-H2S | 18.81| $7.65*10^-6$| 58.2%| 10.89
Taste-H2S-Lactic | 25.26| $6.55*10^-7$| 65.2%| 9.94
Taste-Acetic-H2S-Lactic | 16.22| $3.81*10^-6$| 65.2%$| 10.13  

Out of the three multiple regression models, the H2S/Lactic model performed the best with an adjusted $R^2$ value of 62.6%. The second best model was the model with all 3 explanatory variables which cover an adjusted 61.2% of total variation in Taste. In the two models which included Acetic as a predictor, the p-value for coefficient Acetic was not satistically significant. This result may suggest that the variable Acetic does not add any significant information to the model when paired with H2S or Lactic.  

## Conclusion    
The regression analysis showed that the H2S/Lactic model was the best in predicting Taste. This is because the model has the highest adjusted $R^2$ value of 0.6259 which indicates that the model covers 62.59% of variation in Taste. The resulting regression equation from this model is: $\hat{Taste}$ = -27.6 + 3.95 * $H2S$ + 19.89 * $Lactic$. In addition, the this study suggests that the variable Acetic may not bring as much information to predicting Taste as H2S and Lactic would as exemplified by the last regression model which included all 3 explanatory variables.  

# Section 2: The Details of the Cheese Study  
## Problem 11.53  
```{r, echo=FALSE}
dt <- read.csv("ex11-53cheese.csv", header=TRUE)
n = nrow(dt)
variable <- function(l) {
  obj <- list(x = mean(l), med = median(l), sd = sd(l), iqr = IQR(l))
  attr(obj, 'class') <- 'variable'
  obj
}

taste <- variable(dt$taste)
acetic <- variable(dt$acetic)
h2s <- variable(dt$h2s)
lactic <- variable(dt$lactic)
print("Taste")
stem(dt$taste)
print("Acetic")
stem(dt$acetic)
print("H2S")
stem(dt$h2s)
print("Lactic")
stem(dt$lactic)

par(mfrow=c(2,2))
qqnorm(dt$taste, main="Taste Normal Q-Q Plot")
qqnorm(dt$acetic, main="Acetic Normal Q-Q Plot")
qqnorm(dt$h2s, main="H2s Normal Q-Q Plot")
qqnorm(dt$lactic, main="Lactic Normal Q-Q Plot")
```

Variable           | Taste        | Acetic        | H2s     | Lactic
------------------ | ------------ | --------------| --------| --------
Mean               | `r taste$x`  | `r acetic$x`  |`r h2s$x`| `r lactic$x`
Median             | `r taste$med`| `r acetic$med`| `r h2s$med`| `r lactic$med`
Standard Deviation | `r taste$sd` | `r acetic$sd` |`r h2s$sd`| `r lactic$sd`
IQR                | `r taste$iqr`| `r acetic$iqr`|`r h2s$iqr`| `r lactic$iqr` 

The variables do not have any alarming deviations from normality. Taste and H2s are slightly right-skewed while Acetic seems to have 2 peaks.  

## Problem 11.54 
```{r, echo=FALSE}
par(mfrow=c(2,3))
plot(dt$acetic, dt$taste, xlab="Acetic", ylab="Taste", main="Taste vs Acetic")
plot(dt$h2s, dt$taste, xlab="H2s", ylab="Taste", main="Taste vs H2s")
plot(dt$lactic, dt$taste, xlab="Lactic", ylab="Tastee", main="Taste vs Lactic")
plot(dt$acetic, dt$h2s, xlab="Acetic", ylab="H2s", main="H2s vs Acetic")
plot(dt$acetic, dt$lactic, xlab="Acetic", ylab="Lactic", main="Lactic vs Acetic")
plot(dt$h2s, dt$lactic, xlab="H2s", ylab="Lactic", main="Lactic vs H2s")

taste_acetic = cor(dt$taste, dt$acetic)
taste_h2s = cor(dt$taste, dt$h2s)
taste_lactic = cor(dt$taste, dt$lactic)
acetic_h2s = cor(dt$acetic, dt$h2s)
acetic_lactic = cor(dt$acetic, dt$lactic)
h2s_lactic = cor(dt$h2s, dt$lactic)

# Taste-Acetic
p1 = cor.test(dt$taste, dt$acetic)$p.value
# Taste-H2s
p2 = cor.test(dt$taste, dt$h2s)$p.value
# Taste-Lactic
p3 = cor.test(dt$taste, dt$lactic)$p.value
# Acetic-H2s
p4 = cor.test(dt$acetic, dt$h2s)$p.value
# Acetic-Lactic
p5 = cor.test(dt$acetic, dt$lactic)$p.value
# H2s-Lactic
p6 = cor.test(dt$h2s, dt$lactic)$p.value
```

Pair          | Correlation Coefficient (r)   | P-value
--------------| ----------------------------- | ---------
Taste-Acetic  | `r taste_acetic`              | `r p1`
Taste-H2s     | `r taste_h2s`                 | `r p2`
Taste-Lactic  | `r taste_lactic`              | `r p3`
Acetic-H2s    | `r acetic_h2s`                | `r p4`
Acetic-Lactic | `r acetic_lactic`             | `r p5`
H2s-Lactic    | `r h2s_lactic`                | `r p6`  

All 6 plots show a positive association between each pair of variables.  

## Problem 11.55  
```{r, echo=FALSE}
# Taste: Response, Acetic: explanatory
taste_acetic_lm <- lm(dt$taste ~ dt$acetic)
print(taste_acetic_lm)
summary(taste_acetic_lm)
```
```{r, echo=FALSE}
par(mfrow=c(1,2))
plot(dt$acetic, resid(taste_acetic_lm), main="Residuals vs. Acetic", xlab="Acetic", ylab="Residual")
abline(0,0)

qqnorm(resid(taste_acetic_lm))
qqline(resid(taste_acetic_lm))

par(mfrow=c(1,2))
hist(resid(taste_acetic_lm), main="Histogram of Residuals", xlab="Residual", ylab="Freqeuncy")
plot(dt$acetic, dt$taste, main="Taste vs. Acetic", xlab="Acetic", ylab="Taste")
abline(taste_acetic_lm)

par(mfrow=c(1,2))
plot(dt$h2s, resid(taste_acetic_lm), main="Residuals vs. H2s", xlab="H2s", ylab="Residual")
abline(0,0)
plot(dt$lactic, resid(taste_acetic_lm), main="Residuals vs Lactic", xlab="Lactic", ylab="Residual")
abline(0,0)
```
    
$\hat{Taste}$ = -61.5 + 15.65 * $Acetic$  
The linear model has a p-value of 0.002 which signifies that there is a statistically significant relationship between Taste and Acetic. In addition, the residuals of the taste-acetic linear model seem to have a normal distribution. When the residuals are plotted against H2S and Lactic, they seem to have a positive association. 

## Problem 11.56  
```{r, echo=FALSE}
# Taste: response; H2s: explanatory
taste_h2s_lm <- lm(dt$taste ~ dt$h2s)
print(taste_h2s_lm)
summary(taste_h2s_lm)
```
```{r, echo=FALSE}
par(mfrow=c(1,2))
plot(dt$h2s, resid(taste_h2s_lm), main="Residuals vs. H2S", xlab="H2s", ylab="Residual")
abline(0,0)
qqnorm(resid(taste_h2s_lm))
qqline(resid(taste_h2s_lm))

par(mfrow=c(1,2))
hist(resid(taste_h2s_lm), main="Histogram of Residuals", xlab="Residual")
plot(dt$h2s, dt$taste, main="Taste vs. H2S", xlab="H2S", ylab="Taste")
abline(taste_h2s_lm)

par(mfrow=c(1,2))
plot(dt$acetic, resid(taste_h2s_lm), main="Residuals vs Acetic", xlab="Acetic", ylab="Residual")
abline(0,0)
plot(dt$lactic, resid(taste_h2s_lm), main="Residuals vs Lactic", xlab="Lactic", ylab="Residual")
abline(0,0)
```
  
$\hat{Taste}$ = -9.787 + 5.776 * $H2S$  
The linear model has a p-value of $1.37*10^-6$ which signifies that there is a statistically significant relationship between Taste and H2S. Based on the residual normal Q-Q plot and histogram, the residuals seem to have a normal distribution. When residuals are plotted against Acetic and Lactic, there seems to be larger scatter in the Lactic plot for higher values.  

## Problem 11.57  
```{r, echo=FALSE}
# Taste: response; Lactic: explanatory
taste_lactic_lm <- lm(dt$taste ~ dt$lactic)
print(taste_lactic_lm)
summary(taste_lactic_lm)
```
```{r, echo=FALSE}
par(mfrow=c(1,2))
plot(dt$lactic, resid(taste_lactic_lm), main="Residuals vs. Lactic", xlab="Lactic", ylab="Residual")
abline(0,0)
qqnorm(resid(taste_lactic_lm))
qqline(resid(taste_lactic_lm))

par(mfrow=c(1,2))
hist(resid(taste_lactic_lm), main="Histogram of Residuals", xlab="Residual")
plot(dt$lactic, dt$taste, main="Taste vs. Lactic", xlab="Lactic", ylab="Taste")
abline(taste_lactic_lm)

par(mfrow=c(1,2))
plot(dt$acetic, resid(taste_lactic_lm), main="Residuals vs. Acetic", xlab="Acetic", ylab="Residual")
abline(0,0)
plot(dt$h2s, resid(taste_lactic_lm), main="Residuals vs. H2S", xlab="H2S", ylab="Residual")
abline(0,0)
```
  
$\hat{Taste}$ = -29.86 + 37.72 * $Lactic$  
The p-value of the coefficient is significantly close to 0 which implies that there is a statistically significant relationship between Taste and Lactic. The residuals seem to be normally distributed based on its histogram and normal Q-Q plot. When plotted against the other two chemical variables, the plot seems to be randomly scattered with no real distinguishable pattern.  

## Problem 11.58  
Linear Model | $F$  | $P-value$ | $R^2$  | $s$
------------ | ---- | --------- | ------ | -----
Taste-Acetic | 12.11| 0.002     | 30.2%  | 13.82
Taste-H2S    | 37.29| $1.37*10^-6$| 57.1%| 10.83
Taste-Lactic | 27.55| $1.41*10^-5$| 49.6%| 11.75  

$\hat{Taste}$ = -61.5 + 15.65 * $Acetic$  
$\hat{Taste}$ = -9.787 + 5.776 * $H2S$  
$\hat{Taste}$ = -29.86 + 37.72 * $Lactic$  
The intercepts are different because the linear regression models use different explanatory variables to predict Taste.  

## Problem 11.59  
```{r, echo=FALSE}
# Multiple regression with Acetic and H2S to predict Taste
acetic_h2s_lm <- lm(dt$taste ~ dt$acetic + dt$h2s)
summary(acetic_h2s_lm)
```
  
For the Acetic coefficient, this linear model is not better than the model with only Acetic as an explanatory variable. The p-value in this model for the Acetic coefficient is 0.406 which does not signify a statistically significant relationship between the Acetic variable and Taste in this model. Acetic and H2S have a correlation coefficient of 0.618 and H2S in this model has a statistically significant relationship to Taste. Therefore, the Acetic variable does not add significant information to help us predict Taste if H2S is included due to the two explanatory variables' correlation.  

## Problem 11.60  
```{r, echo=FALSE}
h2s_lactic_lm <- lm(dt$taste ~ dt$h2s + dt$lactic)
summary(h2s_lactic_lm)
```
  
Our multiple regression model using H2S and Lactic to predict Taste is a better predictor for Taste and the model covers 65.2% of the variation in Taste, which is higher than the 57.1% and 49.6% coverage of variation for H2S and Lactic respectively. Additionally, both coefficients in this model has statistically significant p-values (0.002 for H2s and .02 for Lactic). These significant p-values suggest that, in this model, H2S and Lactic have a statistically significant relationship to Taste.  

## Problem 11.61  
```{r, echo=FALSE}
final_lm <- lm(dt$taste ~ dt$acetic + dt$h2s + dt$lactic)
summary(final_lm)

par(mfrow=c(1,2))
hist(resid(final_lm), main="Histogram of Residuals", xlab="Residual")
qqnorm(resid(final_lm))
qqline(resid(final_lm))

par(mfrow=c(1,3))
plot(dt$acetic, resid(final_lm), main="Residuals vs. Acetic", xlab="Acetic", ylab="Residual")
abline(0,0)
plot(dt$h2s, resid(final_lm), main="Residuals vs. H2S", xlab="H2S", ylab="Residual")
abline(0,0)
plot(dt$lactic, resid(final_lm), main="Residuals vs. Lactic", xlab="Lactic", ylab="Residual")
abline(0,0)
```
  
The coefficient of Acetic is not statistically different from 0 as its p-value is 0.94198 (significantly high). However, from the multiple regression model, H2S and Lactic have p-values of 0.004 and 0.031 respectively which signify a statistically signifcant relationship to Taste. Based on its histogram and QQ plot, the residuals seem to be normally distributed and show no significant patterns when plotted with the explanatory variables. The model using H2S and Lactic as predictors for Taste is the best as it covers the most variation in Taste with its adjusteed $R^2$ value of 0.6259. In addition, all of its coefficients are statistically significant within the model.  
